---
title: "Multivariate models"
output: html_document
---

The goals of this lesson are to introduce multivariate ordination analyses.

##Outline
* Overview of ordination methods
* Indirect or Unconstrained Ordination
    - Principle Components Analysis (PCA)
    - Correspondance Anlysis (CA) 
    - Detrended Coresspondance Analysis (DCA)
    - Non-metric Multidimenstional Scaling (NMDS)
* Direct or Constrained Ordination
    - Redundancy Analysis (RDA)
    - Cononical Correspondance Analysis (CCA)
    - Hypothesis Testing
    - Model Comparison

```{r setup, echo=FALSE}
# setup the R enviornment for kniting markdown doc properly
library(knitr)
opts_knit$set(root.dir='../')
```

## Load data and create community matrix
```{r}
library(vegan)
source('./scripts/cleanplot.pca.R')

# load data
trees = read.csv('./data/treedata.csv')
# create a community site x species matrix
comm = tapply(trees$cover, INDEX = list(trees$plotID, trees$spcode), sum)
comm[1:5, 1:5]
# replace the NAs with zeros
comm = ifelse(is.na(comm), 0, comm)
comm[1:5, 1:5]
# visually explore the cover variable between species and sites
sp_sum = apply(comm, 2, sum)
site_sum = apply(comm, 1, sum)
par(mfrow=c(2,2))
hist(sp_sum)
plot(sp_sum[order(sp_sum, decreasing=T)], type='o', col='red', lwd=2,
     xlab='Sp Rank', ylab='Sum Cover')
hist(site_sum)
plot(site_sum[order(site_sum, decreasing=T)], type='o', col='red', lwd=2,
     xlab='Site Rank', ylab='Sum Cover')

## Overview of ordination methods

<insert_table>

## Indirect or Unconstrained Ordination
### Principle Components Analysis (PCA)

tree_pca = rda(comm, scale=TRUE)
plot(tree_pca)
biplot(tree_pca)
cleanplot.pca(tree_pca)


## Reading
* Chapters 5 and 6 of *Numerical Ecology with R*

## Online Docs
* [The Ordination Webpage](http://ordination.okstate.edu/)
    - great for term definitions, layman's explanation of how the methods
    differ, and how ecologists should interpret
* [Vegan: an introduction to ordination](http://cran.r-project.org/web/packages/vegan/vignettes/intro-vegan.pdf)
    - A brief demonstration of an ordination analysis in the R package vegan
* [Multivariate Analysis of Ecological Communities in R: vegan tutorial](http://cc.oulu.fi/~jarioksa/opetus/metodi/vegantutor.pdf)
    - A more thorough demonstation of ordiation in the R package vegan

Table of 
<table cellpadding="0" cellspacing="2">
    <tr>
		<td width="148" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt"><b>Acronym</b></font></font></p>
		</td>
		<td width="150" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt"><b>Name</b>&nbsp;</font></font></p>
		</td>
		<td width="113" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt"><b>type
			of analysis</b></font></font></p>
		</td>
		<td width="124" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt"><b>Algorithm</b>&nbsp;</font></font></p>
		</td>
		<td width="148" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt"><b>R
			function</b></font></font></p>
		</td>
	</tr>
	<tr>
		<td width="97" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt"><i>PCA</i></font></font></p>
		</td>
		<td width="150" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt">Principal
			Components Analysis&nbsp;</font></font></p>
		</td>
		<td width="113" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt">indirect</font></font></p>
		</td>
		<td width="124" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt">Eigenanalysis,
			SVD&nbsp;</font></font></p>
		</td>
		<td width="148" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt">Many</font></font></p>
		</td>
	</tr>
	<tr>
		<td width="97" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt"><i>CA</i></font></font></p>
		</td>
		<td width="150" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt">Correspondence
			Analysis&nbsp;</font></font></p>
		</td>
		<td width="113" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt">indirect</font></font></p>
		</td>
		<td width="124" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt">RA,&nbsp;eigenanalysis,
			SVD&nbsp;</font></font></p>
		</td>
		<td width="148" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt">DECORANA,
			CANOCO, others</font></font></p>
		</td>
	</tr>
	<tr>
		<td width="97" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt"><i>DCA</i></font></font></p>
		</td>
		<td width="150" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt">Detrended
			Correspondence Analysis&nbsp;</font></font></p>
		</td>
		<td width="113" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt">indirect</font></font></p>
		</td>
		<td width="124" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt">RA
			with detrending and rescaling&nbsp;</font></font></p>
		</td>
		<td width="148" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt">DECORANA,
			CANOCO</font></font></p>
		</td>
	</tr>
	<tr>
		<td width="97" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt"><i>CCA&nbsp;</i></font></font></p>
		</td>
		<td width="150" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt">Canonical
			Correspondence Analysis&nbsp;</font></font></p>
		</td>
		<td width="113" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt">direct</font></font></p>
		</td>
		<td width="124" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt">RA
			with regressions,eigenanalysis&nbsp;</font></font></p>
		</td>
		<td width="148" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt">CANOCO</font></font></p>
		</td>
	</tr>
	<tr>
		<td width="97" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt"><i>DCCA</i></font></font></p>
		</td>
		<td width="150" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt">Detrended
			Canonical Correspondence Analysis&nbsp;</font></font></p>
		</td>
		<td width="113" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt">direct</font></font></p>
		</td>
		<td width="124" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt">RA
			with regressions and detrending&nbsp;</font></font></p>
		</td>
		<td width="148" style="border: none; padding: 0in">
			<p><font face="Times New Roman, serif"><font size="3" style="font-size: 12pt">CANOCO</font></font></p>
		</td>
	</tr>
</table>

