---
title: "Multivariate models"
output: html_document
---

The goals of this lesson are to introduce multivariate ordination analyses.

##Outline
* Overview of ordination methods
* Indirect or Unconstrained Ordination
    - Principle Components Analysis (PCA)
    - Correspondance Anlysis (CA) 
    - Detrended Coresspondance Analysis (DCA)
    - Non-metric Multidimenstional Scaling (NMDS)
* Direct or Constrained Ordination
    - Redundancy Analysis (RDA)
    - Cononical Correspondance Analysis (CCA)
    - Hypothesis Testing
    - Model Comparison
    - Variance partitioning


```{r setup, echo=FALSE}
# setup the R enviornment for kniting markdown doc properly
library(knitr)
opts_knit$set(root.dir='../')
```

## Load data and create community matrix
```{r}
library(vegan)
source('./scripts/cleanplot.pca.R')

# load data
trees = read.csv('./data/treedata.csv')
# In class Excersise 1: create a community site x species matrix
comm = tapply(trees$cover, INDEX = list(trees$plotID, trees$spcode), sum)

comm[1:5, 1:5]
# replace the NAs with zeros
comm = ifelse(is.na(comm), 0, comm)
comm[1:5, 1:5]
# visually explore the cover variable between species and sites
sp_sum = apply(comm, 2, sum)
site_sum = apply(comm, 1, sum)
par(mfrow=c(2,2))
hist(sp_sum)
plot(sp_sum[order(sp_sum, decreasing=T)], type='o', col='red', lwd=2,
     xlab='Sp Rank', ylab='Sum Cover')
hist(site_sum)
plot(site_sum[order(site_sum, decreasing=T)], type='o', col='red', lwd=2,
     xlab='Site Rank', ylab='Sum Cover')
```

## Overview of ordination methods

There are generally considered to be two types of ordination. 

1. Indirect or unconstrained ordination in which only a single matrix is analyzed
2. Direct or constrained ordination in which one matrix is used to explain the 
variance of another matrix. 

Today we will discuss both types. In general ordination is considered be an ideal
approach 

[Ordination methods table](http://dmcglinn.github.io/quant_methods/lessons/ordination_methods_table)

## Indirect or Unconstrained Ordination
### Principle Components Analysis (PCA)

tree_pca = rda(comm, scale=TRUE)
plot(tree_pca)
biplot(tree_pca)
cleanplot.pca(tree_pca)


## Reading
* Chapters 5 and 6 of *Numerical Ecology with R*

## Online Docs
* [The Ordination Webpage](http://ordination.okstate.edu/)
    - great for term definitions, layman's explanation of how the methods
    differ, and how ecologists should interpret
* [Vegan: an introduction to ordination](http://cran.r-project.org/web/packages/vegan/vignettes/intro-vegan.pdf)
    - A brief demonstration of an ordination analysis in the R package vegan
* [Multivariate Analysis of Ecological Communities in R: vegan tutorial](http://cc.oulu.fi/~jarioksa/opetus/metodi/vegantutor.pdf)
    - A more thorough demonstation of ordiation in the R package vegan
