---
title: "R Introduction"
author: "Dan McGlinn"
date: "Sunday, January 18, 2015"
output: html_document
---

The purpose of this lesson is to introduce students to the R programming 
environment for the first time. The lesson builds off the 
[Software Carpentry](http://software-carpentry.org/) Lesson developed here: 
<http://swcarpentry.github.io/r-novice-inflammation/01-starting-with-data/index.html>

## Readings
* Chapters 1-7 of *The R Book* (1st ed) by Crawley
* Chapters 1-4 of *MASS* (4th ed)by Venables and Ripley

## Lesson Outline
* <a href="#arith">Arithmetic</a>
* <a href="#logic">Logical operations</a>
* <a href="#vars">Variable assignment</a>
* <a href="#read">Reading in data</a>
* <a href="#help">Using the help</a>
* <a href="#data">Examine data</a>
* <a href="#subset">Subsetting the data</a>
* <a href="#summary">Summary statistics</a>
* <a href="#aggr">Aggregate across rows or columns</a>
* <a href="#plot">Plot data

```{r setup, echo=FALSE}
  library(knitr)
  opts_knit$set(root.dir='../')
```

## <a href="#arith" name="arith">#</a> Arithmetic

```{r}
3 + 4       # summation

3 * 4       # multiplication

3 / 4       # division

3^4         # exponents

log(3)      # log base e

log(3, 10)  # log base 10

log10(3)    # log base 10

exp(log(3)) # e
```

## <a href="#logic" name="logic">#</a>Logical operations

```{r logical operations}
?logical 

3 > 4        # greater than

3 < 4        # less than

3 >= 4       # greater than or equal to

3 <= 4       # less than or equal to 

3 != 4       # not equal to 

3 == 4       # equal to

TRUE         # True
T            # True
TRUE == 1    # True is set to one in R

FALSE        # False
F            # False
FALSE == 0   # False is set to zero in R

## operations 
# ! reverses a logical
!FALSE

# & can be used as an "and" statement
T & T
T & F

# | can be used as an "or" statement
T | T
T | F

## logical algebra
# T is treated as a 1 
# F is treated as a 0
T + T + F    # what would this equal?

## useful functions
# any() and all()
any(c(T, F))
all(c(T, F))
```

## <a href="#vars" name="vars">#</a>Variable assignment 
You can use "<-" or "=" to assign a value to a variable
```{r variable assignment}
weight_kg <- 55

# print the value of the variable by simply calling its name
weight_kg

# weight in pounds:
2.2 * weight_kg

weight_kg <- 57.5

# weight in kilograms is now
weight_kg

weight_lb <- 2.2 * weight_kg
# weight in kg...
weight_kg
# ...and in pounds
weight_lb

weight_kg <- 100.0
# weight in kg now...
weight_kg
# ...and in weight pounds still
weight_lb
```

Coming up with good object and file names can be difficult, but there 
are two general rules that can help guide you: 
   
    1) be descriptive
    2) don't make names you must type a lot too long

So for something like a file name which you'll only type probably once at read and
write you should use a long descriptive name, but for objects in your R code you
need to consider typeability and readability when designing the name. A long name
like root_rhiz_prod_total_mm is very clear but is a pain to read and worse to 
type. R has a built-in name completion system but this doesn't completely 
remove the burden on you for using long object names. 

## <a href="#read" name="read">#</a>Reading in data

First check what your working directory is:
```{r}
getwd()
```

because I have setup a Project in the quant_methods folder I can make my
path relative to this location.

let's read in the datafile `inflammation-01.csv` which is located in the
directory: `./quant_methods/data)` where the `.` indicates the directory
location in which the directory `quant_methods` is stored in. The usage of 
the `.` is a shorthand way to create relative paths. 
Because my working directory is already set to: ``r normalizePath('.')``
I can shorten the path to `./data/inflammation-01.csv` where again `.` 
indicates my current working directory path. 

```{r read in data}
dat <- read.csv(file = "./data/inflammation-01.csv", header = FALSE)
```

Alternatively and a bit more cleanly we can take advantage of the ability to 
supply the function `read.csv` with a url as such

```{r url import}
dat <- read.csv('http://dmcglinn.github.io/quant_methods/data/inflammation-01.csv',
               header=F)
```

## <a href="#help" name="help">#</a>Using the help

```{r get help, eval=FALSE}
# above we used the function "read.csv" to find out more about this function see
?read.csv 
# or equivalently 
help(read.csv) 
# to do a fuzzy help search use
help.search('read') 
help.search('csv')
```

## <a href="#data" name="data">#</a>Examine data

```{r examine data}
# visual summary of first 6 rows
head(dat)
# visual summary of last 6 rows
tail(dat)

# what kind of object is dat
class(dat)

# what are the dimensions of dat
dim(dat)
```

You may notice that the data did not have column names and R auto assigned the
columns the names V1, V2, V3, and so on. In this dataset, each column represent
different times. We can assign column names using the function `names`
```{r setting column names}
names(dat)
names(dat) <- paste("day", 1:ncol(dat), sep='')
names(dat)
```

Above the function `paste` was used to construct text strings that combined the
word "patient" with a given index in this case from 1 to the total number of 
columns in the object `dat`. By default the function `paste` inserts a space 
between strings that you wish to paste together, I've set the `sep` argument 
to `''` to ensure that no space is inserted (see also `?paste0`)

## <a href="#subset" name="subset">#</a>Subsetting the data

There are a variety of ways to subset data in data.frames. This section
demonstrates how to subset data using indices. 
```{r subset index}
# first value in dat
dat[1, 1]

# middle value in dat
dat[30, 20]

# chunk of data in dat
dat[1:4, 1:10]

# select specific rows and columns
dat[c(3, 8, 37, 56), c(10, 14, 29)]

# all columns from row 5
dat[5, ]
# all rows from column 16
dat[ , 16]
dat[1:nrow(dat), 16]

#first 5 rows and all columns except 16
dat[1:5, -16]
```

An alternative way to carry out subsetting is to reference specific column 
names or to use the `subset` function

```{r subset names}
# here to avoid printing too much information to the screen I'll just focus on
# on the first 5 rows of each subset
dat$patient10[1:5]
dat[1:5 , 'day10']
dat[1:5 , c('day10', 'day15')]
# notice that the following would give and error
#dat[ , -c('patient3')]
# but that the following would accompish the intended goal of dropping patient 3
dat[1:5 , -3]

#let's try using the subset function
# only data for day 3
subset(dat, select = day3)[1:5, ]
# data on all days but 3
subset(dat, select = -day3)[1:5, ]
# data only on day 3 when inflammation in day 1 is equal to 0
subset(dat, subset = day1 == 0, select = day3)[1:5, ]
```

## <a href="#summary" name="summary">#</a>Summary statistics
```{r}
# first row, all of the columns
patient_1 <- dat[1, ]
# max inflammation for patient 1
max(patient_1)

# max inflammation for patient 2
max(dat[2, ])

# minimum inflammation on day 7
min(dat[ , 7])

# mean inflammation on day 7
mean(dat[ , 7])
# median inflammation on day 7
median(dat[ , 7])
# standard deviation of inflammation on day 7
sd(dat[ , 7])

summary(dat[ , 7])
```

## <a href="#aggr" name="aggr">#</a>Aggregate across rows or columns

To obtain the average inflammation of each patient we will need to
calculate the mean of all of the rows (`MARGIN = 1`) of the data frame.

```{r}
avg_patient_inflammation <- apply(dat, 1, mean)
```

And to obtain the average inflammation of each day we will need to calculate
the mean of all of the columns (`MARGIN = 2`) of the data frame.

```{r}
avg_day_inflammation <- apply(dat, 2, mean)
```

We can change the function "mean" for other functions such as "sd" which
calculates the standard deviation

```{r}
# standard deviation of day
sd_day_inflammation <- apply(dat, 2, sd)

# standard deviation of patients
sd_patient_inflammation <- apply(dat, 1, sd)
```
## <a href="#plot" name="plot">#</a>Plot data

```{r plot help, eval=FALSE}
# use the function plot() to plot data
?plot
```

This help file provides a long list of potential arguments and examples
at a minimum you must provide a single quantitative variable, for example:

```{r default plot}
plot(avg_day_inflammation)
```

Notice how R fills in lots of pieces of missing information automatically. 
specifically it assumes that the independent variable is simply an index from
1 to the length of the object in this case avg_day_inflammation. A safer more 
clear way to accomplish the same plot is to use the following:

```{r}
plot(1:length(avg_day_inflammation), avg_day_inflammation, xlab='day', 
    ylab='inflammation')
```

This makes it clearer that the x-variable is simply an index from 1 to the
length of avg_day_inflammation, and it makes the x and y axis labels more 
sensical. 

To output multi-panel plots use for example
```{r}
par(mfrow=c(2,1))
# which will create a single plotting row with two columns
plot(1:length(avg_day_inflammation), avg_day_inflammation, xlab='day', 
     ylab='inflammation')
plot(1:length(avg_patient_inflammation), avg_patient_inflammation,
     xlab='patient identity', ylab='inflammation')
```

To output the figure to file you can use Rstudio's GUI features or you can use
the command line which is what I recommend so that the code is fully 
reproducible:

```{r make pdf, eval=FALSE}
pdf('./lessons/inflammation_fig1.pdf')
par(mfrow = c(2,1))
plot(1:length(avg_day_inflammation), avg_day_inflammation, xlab='Day',
     ylab='Inflammation', frame.plot=F, col='magenta', pch=2, cex=2)
plot(1:length(avg_patient_inflammation), avg_patient_inflammation,
     xlab='patient identity', ylab='inflammation', col='dodgerblue')
dev.off()
```

