---
title: "Simulations in R"
output: html_document
---

R provides an excellent enviornment for carrying out numerical experiments or
simulations. These can be very helpful for increasing mathematical competancy, 
testing hypotheses, assessing statistical power, and generating theoretical 
expectations. 

We've already experimented a bit with simulations in this course:

1. In the univariate lesson we [simulated variables that we then used in a
regression analysis](http://dmcglinn.github.io/quant_methods/lessons/univariate_models.html#sim).
2. In the multivariate lesson we [permuted rows of our matrix to test if our
constrained ordination model fit the data better than we would expect due to
chance](http://dmcglinn.github.io/quant_methods/lessons/multivariate_models.html).
3. In the the spatial lesson we [permuted the spatial coordinates of samples to 
test if the spatial correlation was larger than we would expect due to chance](http://dmcglinn.github.io/quant_methods/lessons/spatial_models.html)

In the first example listed above the simulation was designed to inform our 
understanding of regression and to examine how sensative it was to violations of
its assumptions. In examples 2 and 3 listed above the role of the simluation was
to generate a null model that an observed pattern could be tested against. 

Here I present an example of a simulation of a theorecial model to try to gain
an understanding of this model and of a broader concept known as Chaos theory.

I was inspired to code this in R after watching this video by 
[Veritasium - This equation will change how you see the world](https://www.youtube.com/watch?v=ovJcsL7vyrk)

# Chaotic logistic population equlibria 

First we are going to define a simple theorecial model of population growth that
has a negative feedback. In other words as the population grows its rate of 
increase decreases as it approches a carrying capacity for example. This kind of
growth model is known as a logistic growth model and it constrats from the 
exponential model of growth in which the population continues to grow to infinity.
The logistic model can be presented mathematically as:

$$\dfrac{dN}{dt} = rN(1-N)$$
where *N* is population size, *t* is time, and *r* is the intrinic rate of
population growth rate on a per captia basis. In this formulation, the 
carrying capacity of the population is implicitly defined as equal to 1. 

Let's code this simple model up in R and then examine its behavior with a simulation


```{r}
# simple model of logistic growth
dNt <- function(r, N) r * N * (1 - N)

# iterate growth through time
Nt <- function(r, N, t) {
  for (i in 1:(t - 1)) {
    # population at next time step is population at current time + pop growth
    N[i + 1] <- N[i] + dNt(r, N[i])
  }
  N
}
```

Now let's examine what happens in this model if we run it through time for 
different starting abundance values.

```{r}
t <- 100
r <- 0.1
# lets consider 4 different starting abundances (i.e., N(t=0) values)
Nt0 = c(0.1, 0.5, 1.5, 2)

par(mfrow=c(2,2))
for (i in seq_along(Nt0)) {
  plot(1:t, Nt(r, Nt0[i], t), type = 'l', xlab = 'time', ylab = 'Population size',
    main = paste('N(t=0) =', Nt0[i]), ylim =c(0, 2))
  abline(h = 1, lty = 2, col='grey')
}

```

So what we learn from this is if you start below the carrying capacity (grey 
dashed line) your population will increase to it, and if you start above 
carrying capacity your population will decrease to it. We also learn that the 
rate of increase or decrease depends on how far from the equilibrium you are at
the beginning of the time series. 

Ok so all is well in the world of logistic population growth - given enough time
all of the populations eventually hit carrying capacity 
and do not change from that point. Therefore, if we wanted to predict 
long-term (i.e., equilibrium) abundance then this model would seem to suggest
that should be 1 for all of these populations because they all have the same 
carrying capacity of 1. 

To examine if this is actually true let's examine the other parameter of the model
that we can adjust *r* of the rate of population growth. 

Question: do all of the populations go to the same expected population 
equilibrium (i.e., the carrying capacity or 1 in this spectific case)? 

To test this we will examine a range of *r*-values from 0.01 to 3, we'll set the
starting population size to 0.5 but where you actually start N out ends up not
making a difference for addressing this specific question. The approach laid out
below we will assume that equilibrium is hit at the end of the first of the time
series. This is would seem to be a pretty reasonable assumption given what we
learned above and we'll run the simluation 10 times longer (1000 time steps) to 
make extra sure that the model should have achieved equilibrium. 

```{r}

# set starting conditions and amount of time 
t <- 1000
r <- seq(0.01, 3, .01)
Nt0 <- 0.5
# compute the population sizes across the times
e <- sapply(r, function(r) Nt(r, Nt0, t))

# only use 2nd half of times presuming those will be at equilibrium
thalf <- round(t/2)
e <- e[thalf:t, ]
t <- nrow(e)
maxE <- max(as.vector(e))

# plot simulation results

ptsize = 0.25
plot(rep(r[1], t), e[ , 1], ylim = c(0, maxE), xlim = range(r), 
     cex = ptsize, xlab = 'Population growth rate (r)', 
     ylab = 'Equilibrium abundance')
for (i in seq_along(r)) {
  points(rep(r[i], t), e[ , i], cex = ptsize)  
}
abline(h = 1, col='grey', lty=2)
```

What we see above was quite unexpected to the ecologist Robert May who first
discovered this phenomena ([May 1976](https://www.nature.com/articles/261459a0)). 
Essentially what this figure tells us is that for population growth rates above
2 that the population does not necessarily stay at the equilibrium where we expect it
to (at the dashed grey line, 1). Instead equilibrium abundances bifurcated into 
different values and further bifurcate until essentially any abundance value is 
possible at r values greater than 2.6 or so. 

The important thing to remember about the result above is that:

1. this is a purely deterministic model - i.e., no error so this complex outcome 
was generated only by the negative feedback loop in the model
2. this is a very simple model so this is not due to an exceptional complex
internal dynamics of the model
3. although equilibrium abundance is unpredicatable at higher r values the overall
pattern of the bifurcation is predicable. It is what folks call a "strange attractor".

As might be imagined this unpredictable behavior of one of the simplist and model
canonical model of population growth shook ecology to its core. The reverberations
of which are still being felt over 40 years later. 
